
\subsection{Independence}

\ssn{Definition: independence} \hfill
\tcb
Two events $A,B \subseteq S$ are \ul{independent} if
 \[
   \PP(A \cap B) = \PP(A) \PP(B). 
 \]
\etcb

\noindent
The event $A \cap B$ is the intersection of the two events: it occurs precisely when both $A$ and $B$ occur. 

We say two events are ``dependent'' if they are not independent. 

We will say more about independence next week.  Two events being independent means that knowing that one has occurred does not change the probability that the other occurs.  
\end{n}

\ssn{Examples}
\begin{itemize}
    \item Roll a D6 twice. Let $A$ be the event that the first roll is a `6' and let $B$ be the event that the second roll is a `6'. Then $\PP(A) = \PP(B) = 1/6$. Then $A \cap B$ is rolling `double-6' which has probability $1/36$. So $A$ and $B$ are independent.  We were using this independence in our calculations with the probability trees above. 
    \item Roll a D6 twice. Let $A$ be the event that the sum of the rolls is $5$ and let $B$ be the event that at least one `1' appears.     Then $A=\{ (1,4), (4,1),(2,3),(3,2) \} $ and so $\PP(A) = 4/36$.   Counting we find $11$ possible rolls contain at least one `1' and so $\PP(B) = 11/36$.    The event $A \cap B = \{ (1,4), (4,1)$ and so $\PP(A \cap B) = 2/36$. Checking 
     \[
        \frac2{36} \not = \frac{4}{36} \, \frac{11}{36}
     \]
     and so $A$ and $B$ are \emph{not} independent. 
      \begin{itemize}
          \item Looking at this from another angle, if we know that the sum is 5, then there are only four possibilities and two of them contain a `1':  the probability now of there being a `1' present is $1/2$. 
      \end{itemize}
\end{itemize}
It is worth noting at this stage that the very last item above is usually expressed in the form 
 \[
  \text{The \ul{conditional probability} of $B$ given $A$ is equal to $1/2$} 
 \]
which we write with the notation 
\[
   \PP( B \st A) = \frac12 
\]
where the vertical bar is read as ``given''. 
We will take up conditional probability further next week. 
\end{n}
