
\subsection{Inclusion-Exclusion Principle}

\ssn{Inclusion-exclusion principle (for two sets)}
  Let $A,B$ be two events.  Then 
  \[
  \PP( A \cup B ) = \PP(A) + \PP(B) - \PP(A \cap B). 
  \]
\begin{proof}
 Consider the following three decompositions into disjoint unions. 
 \begin{eqnarray*}
     A & = & ( A\cap B )  \cup (A \setminus B) \\
     B & =&  ( A\cap B )  \cup (B \setminus A) \\
     A \cup B &=& (A\cap B) \cup  (A \setminus B) \cup (B \setminus A) . 
 \end{eqnarray*}
 Since the unions are disjoint on the right-hand sides the probabilities add. 
  \begin{eqnarray*}
     \PP(A) & =&  \PP( A\cap B )  + \PP( A \setminus B) \\
     \PP(B) & = & \PP( A\cap B )  + \PP(B \setminus A) \\
    \PP( A \cup B) &=& \PP (A\cap B) + \PP(A \setminus B) +\PP(B \setminus A) . 
 \end{eqnarray*}
Subtracting the first two equations from the third we obtain the result. 
\end{proof}
\end{n}



\ssn{Example} 
I choose a natural number $n$ in the interval $[1,100]$ uniformly randomly. What is the probability that $n$ is divisible by 2 or by 5?
\begin{enumerate}
    \item Our ``or'' just above is, as always in mathematics is ``inclusive'': we include the possibility that the number is divisible by both. 
    \item Recall that ``uniformly randomly'' just means all options are equally likely.
    \item Let $A$ be the event that I choose a number divisible by two. There are fifty such numbers and so $\PP(A) = 50/100$. Similarly, let $B$ be the event that I choose a number divisible by five. There are twenty such numbers and so $\PP(B) = 20/100$. 
    \item Then $A \cap B$ is the event that $n$ is divisible by both two and five; equivalently, $n$ is divisible by ten. So $\PP(A \cap B) = 10/100$. 
    \item The probability we want is $\PP(A \cup B)$ which by inclusion-exclusion is 
    \[
     \PP(A \cup B) = \PP(A) + \PP(B) - \PP(A \cap B) = 
      \frac{50}{100} + \frac{20}{100} - \frac{10}{100} = \frac35. 
    \]
\end{enumerate}
\end{n}

\sse{}
I choose a natural number $n$ in the interval $[1,400]$ uniformly randomly. What is the probability that $n$ is divisible by 4 or by 10?  (Take care: the answer is \emph{not} $\frac{13}{40}$.) 
\end{e} 

\sss
A number is divisible by both 4 and 10 if and only if it is divisible by their ``least common multiple'' which is 20.  The answer is $\frac{3}{10}$. 
\end{s}

\sse{} 
I have three red cards numbered $1,2,3$ and three blue cards numbered $1,2,3$.  I shuffle all six cards together thoroughly (so that all possible orderings of the six are equally likely). I then take the top 2 cards from the pile. 

Let $A$ be the event that the two cards are the same colour and let $B$ be the event that at least one of the cards is a `3'. Consider the sample space $S$ for this experiment to be all subsets $\{ x,y \}$ where $x$ and $y$ are different elements of the set $\{ r1, r2, r3, b1, b2, b3\}$. Compute the probabilities of the events
 \[
   A,\quad B,\quad A^c,\quad A \cup B,\quad A \cap B,\quad A \setminus B. 
 \]
(Hint: Your sample space should have 15 elements.) 
\end{e}



